\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}

In this thesis various data-oblivious sorting algorithms are discussed with a focus on practical and concise descriptions promoting ease of implementation.

Furthermore, general performance concerns of data-oblivious  sorting algorithms are explored, along with demonstrations of both SSE and CUDA implementation variants of major data-oblivious sorting algorithms. This is further expanded by presenting a technique for automatic vectorization of general sorting networks.

Finally, several data-oblivious sorting algorithms are compared experimentally based on several parameters, many of which have a direct and measurable effect on real-life performance of the aforementioned algorithms. Along with these findings, we also present actual performance measurements for optimization techniques applied to the data-oblivious sorting algorithms, including both SSE, CUDA and OpenMP, alongside algorithm-specific optimizations. 

\chapter*{Resum\'e}
\addcontentsline{toc}{chapter}{Resum\'e}


I dette speciale diskuteres forskellige data-oblivious sorteringsalgoritmer, med fokus på korte, praktiske beskrivelser for at fremme letheden ved deres implementation.

Derudover udforskes generelle performance-problemer for data-oblivious sorteringsalgoritmer sammen med præsentationer af SSE og CUDA implementationer af vigtige data-oblivious sorteringsalgoritmer.
Dette udvides yderligere med præsentationen af en teknik til automatisk vektorisering af generelle sorteringsnetværk.

Endeligt sammenlignes flere data-oblivious sorteringsalgoritmer baseret på varierende parametre, hvoraf mange har en direkte og målbar effekt på den praktiske udførselstid. Sammen med disse resultater præsenteres også målinger for forskellige optimeringsteknikker for data-oblivious sorteringsalgoritmer inklusiv SSE, CUDA og OpenMP, sammen med algoritmespecifikke optimeringer.

